---
title: "Draft Vignette"
output: html_document
---

I have used several packages 

## Basics

```{r}

library(epldata)

data(package="epldata")
## not best
        

```

Let's look at one of them

```{r}
library(tidyverse) # overkill but easier add suppress warning s o block

glimpse(players)

```


Let's get say percentage distribution of players born from 1990 by birth country 

```{r}

players %>% 
  filter(birth_date>"1989-12-31") %>% 
  group_by(birth_country) %>% 
  tally() %>% 
  mutate(pc=round(100*n/sum(n),2)) %>% 
  arrange(desc(pc))


```

Predictably England dominates but Scotland only scrapes into the top 10

## Combining tables

Much more commonly you will need to combine tables

Which player has scored the most for each team

```{r}

df <- players %>% 
  #filter(player_id!="OWNGOAL") %>% 
  left_join(player_team) %>% #7080 7107 - may be some that never played
  left_join(player_game) %>% 
  right_join(goals) %>% 
  mutate(name=paste(first_name,last_name)) %>% 
  group_by(player_id,name,team) %>% 
  tally() %>% 
  arrange(desc(n)) %>% 
  group_by(team) %>% 
  slice(1) %>% 
  ungroup() %>% 
  filter(!(is.na(team))) %>% ## still to explain
  select(team,name,goals=n)
  
  

df


```


## Derived tables

The above example included quite a few joins which you may not wish to do 
for every analysis
For instance you might want to have the standings for a team at any point of a season

```{r}

## goals by team for individual match

goals_by_team <- game_team %>% 
  left_join(player_game) %>% 
  right_join(goals) %>% 
  group_by(team,team_game_id,game_id) %>% 
  tally() %>% 
  right_join(game_team) %>% 
  mutate(GF=ifelse(is.na(n),0,n)) %>% 
  select(-c(venue,n))

goals_by_team

```

So we now have the goals scored by each team. The next step is to mach theis table with itself to obtain goals against but only by the game_id

```{r}
df <-goals_by_team %>% 
  inner_join(goals_by_team,by="game_id") 

head(df)
```

We have duplication and wish to remove all those where team.x= team.y. as well as tidy up column names and calculate the points accrued for each match

```{r}

df <- df %>% 
  filter(team.x!=team.y) %>% 
  select(team=team.x,team_game_id=team_game_id.x,game_id,GF=GF.x,opponents=team.y,GA=GF.y) %>% 
  mutate(points=case_when(
       GF >GA  ~ 3,
       GF==GA ~ 1,
       GF<GA ~ 0
    
  ))
```

For the standings at any time, we need to add the date in order and split the results into seasons


```{r}

years <- c(1992:2018)

    
library(lubridate)
df_next <- df %>% 
  left_join(game) %>% 
  mutate(year=year(game_date),month=month(game_date)) %>% 
  mutate(season= case_when(
    month<=7 ~ paste(year-1,year,sep="/"),
    month>7 ~ paste(year,year+1,sep="/")
         )
  ) %>% 
  arrange(game_date) %>% 
  group_by(season,team) %>% 
  mutate(year_game_order=row_number())
    
  df_next
```

We can now create a standings data.frame for each round of matches based on points, Goal difference , and Goals For

```{r}

standings <- df_next %>% 
  select(team,season,game_date,year_game_order,GF,GA,points) %>% 
  group_by(team,season,year_game_order) %>% 
  mutate(cum_points=cumsum(points),cum_GF=cumsum(GF),cum_GA=cumsum(GA),cum_GD=cum_GF-cum_GA) %>% 
  group_by(season) %>% 
  arrange(desc(cum_points),desc(cum_GD),desc(cum_GF),team) %>% 
  mutate(position=row_number()) %>% 
  select(season,team,round=year_game_order,position,GF=cum_GF,GA=cum_GA,GD=cum_GD,points=cum_points)

standings

```

It is then a simple matter to create a function to get a table for any round of any year. e.g after 20 games in 1994/1995

```{r}

table_year_round <- function(x,y){
  standings %>% 
    filter(season==x,round==y)
}

table_year_round("1994/1995",20)

```

Obviously you can vary what is in the standings table

and there may be many other derived tables you want to create and save
