---
title: "epldata - visualizations"
#author: "Andrew Clark"
#date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

We have previously covered basic usage of datasets within the package including combining them to produce answers to questions and creating derived tables.

We will now look at more interesting output in the form of

 * Tables
 * Charts
 * Maps
 
 This package is particularly suited to the first two options
 
 You will need the standings data.frame created in the first vignette so if it is not in the environment
 either load a saved version or re-run the code
 
 ### Table
 
 I tend to use the `DT` package, but there are alternatives if you have another preference
 
 Let's use the match_summary_full dataframe to calculate each team's head to head record.  Over and above the current data, we need to create and sum the results
 




```{r }

match_summary_full %>% 
  ungroup() %>%  #match_summary_full is grouped tbl_df
  #filter(team=="Arsenal",opponents=="Chelsea") %>% 
  group_by(team,opponents) %>% 
 # select(GF,GA) %>% 
  mutate(result = case_when(
    GF > GA ~ "W", #win
    GF == GA ~ "D", #draw/tie
    GF < GA ~ "L" # loss
  )) %>% 
  select(team,opponents,result,GF,GA,points) %>%
  mutate(yesno = 1) %>%
  distinct %>%
  spread(result, yesno, fill = 0) %>% 
  summarize(P=n(),W=sum(W),D=sum(D),L=sum(L),ppg=round(sum(points)/P,2))%>%
  arrange(desc(ppg)) %>% 
   DT::datatable(class='compact stripe hover row-border order-column',rownames=FALSE,options= list(paging = TRUE, searching = TRUE,info=FALSE))



```



```{r}




```



```{r}


```



```{r}


```



```{r}

```



```{r}


```




```{r}


```



```{r}



```

