---
title: "epldata - visualizations"
#author: "Andrew Clark"
#date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE }
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

We have previously covered basic usage of datasets within the package including combining them to produce answers to questions and creating derived tables.

We will next look at more interesting output in the form of

 * Tables
 * Charts
 * Maps
 
 This package is particularly suited to the first two options though there is some geographic data to play around with
 
 You will need the the data.frames created earlier  so if it they are not in your environment  either load a saved version or re-run the code
 
 ### Tables
 
 I tend to use the `DT` package, but there are alternatives if you have another preference
 
 Let's use the match_summary_full dataframe to calculate each team's head to head record.  Over and above the current data, we need to create and sum the results
 




```{r }

match_summary_full %>% 
  ungroup() %>%  #match_summary_full is grouped tbl_df
  #filter(team=="Arsenal",opponents=="Chelsea") %>% 
  group_by(team,opponents) %>% 
 # select(GF,GA) %>% 
  mutate(result = case_when(
    GF > GA ~ "W", #win
    GF == GA ~ "D", #draw/tie
    GF < GA ~ "L" # loss
  )) %>% 
  select(team,opponents,result,GF,GA,points) %>%
  mutate(yesno = 1) %>%
  distinct %>%
  spread(result, yesno, fill = 0) %>% 
  summarize(P=n(),W=sum(W),D=sum(D),L=sum(L),ppg=round(sum(points)/P,2))%>%
  arrange(desc(ppg)) %>% 
   DT::datatable(class='compact stripe hover row-border order-column',rownames=FALSE,options= list(paging = TRUE, searching = TRUE,info=FALSE))



```

This provides a sortable, searchable table

---

### Charts

Let's turn attention to players. Firstly I will create a data.frame
for the goals and assists for a specified player

For ease of use below, I have created it as a function and
slipped in an example player_id


```{r message = FALSE}

player_game_data <- function(player) {
# collect goal information for specific player
df_goals <- players %>% 
  left_join(player_team) %>% 
  left_join(player_game) %>% 
  left_join(goals) %>% 
  filter(start==TRUE|time_on>0) %>% 
  select(player_id,last_name,player_game_id,goal_id,team_game_id) %>% 
  mutate(goal=ifelse(!is.na(goal_id),1,0)) %>% 
  group_by(player_id,last_name,team_game_id) %>% 
  summarize(tot_goals=sum(goal)) %>% 
   filter(player_id==player)

# likewise with assists
df_assists <- players %>% 
  left_join(player_team) %>% 
  left_join(player_game) %>% 
  left_join(assists) %>% 
  filter(start==TRUE|time_on>0) %>% 
  select(player_id,last_name,team_game_id,assist_id,player_game_id) %>% 
  mutate(assist=ifelse(!is.na(assist_id),1,0)) %>% 
  group_by(player_id,last_name,team_game_id) %>% 
  summarize(tot_assists=sum(assist)) %>% 
   filter(player_id==player)

# combine
df_all <- df_goals %>% 
  inner_join(df_assists) %>% 
# create a game order  
  left_join(game_team) %>%
  left_join(game) %>% 
  arrange(game_date) %>% 
  mutate(player_game_order=row_number()) %>% 
   ungroup() %>% #removes unwanted name and PLAYERID
  select(player_game_order,tot_goals,tot_assists) %>% 
  # gather into narrow format for plotting
  gather(category,count,-player_game_order) 


}

player_df <-player_game_data("SALAHM")
head(player_df)



```


_You can see why you might want to create a derived player table first if you want to do varied detailed analyses particularly where the raw data is only updated annually .saves time and enhances user interactivity experience_

Now just choose your plotting package of choice to display the data. I will use plotly as this allows for ease of infoactivity with feature susch as panning/zooming, hover tooltips etc

```{r message = FALSE}
library(plotly)

player_df %>%
  plot_ly(x=~player_game_order, y= ~count) %>%
  add_bars(color= ~category, colors=c("red","blue")) %>%
  layout(barmode="stack")

```

Lots of customization is available within the package. 

---

## Interactivity

Lets use the data to create some interactive output

### Shiny

Lets say we use the match_summary_full data to plot a histogram of the goals 
scored by a team in the Premier League


```{r}

library(shiny)
library(glue)

shinyApp(
  
 
  
  ui = fluidPage(
    
    ## calculate an ordered vector of teams to  select from
     teams <- match_summary_full %>%
      pull(team) %>%
      unique() %>% 
      sort(),
     
    selectInput("team", "Select Team:", teams),
    
    plotlyOutput("goals_for")
  ),
  server = function(input, output) {
    
    output$goals_for <- renderPlotly({
      
      match_summary_full %>%
        filter(team == input$team) %>%
        plot_ly %>%
        add_histogram(x =  ~ GF) %>%
        layout(title = glue("Distribution of Goals scored by {input$team}"))
      
    })
  }
)


```


### Crosstalk

This is an alternative  -- dont want to go to server interaction by brushing

#### Filtering


```{r warning = FALSE}
library(crosstalk)

msf  <- SharedData$new(match_summary_full)
bscols(
  widths = c(12), # forces components into rows
filter_select(id="team",label="Select a Team",sharedData=msf, group =  ~team, multiple = FALSE),
plot_ly(msf, x = ~GF, showlegend = FALSE) %>% 
    add_histogram(color = ~team, colors = "red")
)
```

## Animation
see plotlyfunctionsextended.RMD which I think I did as blog post

We can use the  standings dataset prepared earlier




Let's look at how arch-rivals, Brighton and Crystal Palace, fared last season

```{r animation}

# function to add cumulative line
# courtesy Carson Sievert

accumulate_by <- function(dat, var) {
  var <- lazyeval::f_eval(var, dat)
  lvls <- plotly:::getLevels(var)
  dats <- lapply(seq_along(lvls), function(x) {
    cbind(dat[var %in% lvls[seq(1, x)], ], frame = lvls[[x]])
  })
  dplyr::bind_rows(dats)
}

# select team(s) to display 
teams <- c("Brighton","Crystal P")

# add function to base data and year of interest
df <- standings %>% 
  filter(season=="2017/2018"&team %in% teams) %>% 
  accumulate_by(~round)

# static plot - scatter plot- uncolored
base <- df %>% 
  plot_ly(x=~round,y=~position) %>% 
layout(
    xaxis=list(title="Games Played"),
   yaxis=list(title="League Standing",range=c(20.5,0.5))
  ) %>% 
  config(displayModeBar = F,showLink = F) 


# add animation options and color-blind safe colors
 base %>%
   add_lines(color = ~team, colors="Set2", frame = ~frame, ids = ~team) %>%
  animation_opts(500, easing = "linear",mode='immediate') %>%
  animation_button(
    x = 1, xanchor = "right", y = 0, yanchor = "middle", font = list(color="red"), bgcolor="lightblue"
  ) %>%
  animation_slider(
    currentvalue = list(prefix = "Game ")
  ) 

```

Brighton, a promoted club, were expected to be struggle but Crystal Palace spent
more of the season in danger of relegation. In the end, they both survived relegation by placing higher than 18th
